---
title: "Practicum I V2 CS5200"
authors: Rupert Simpson, Paula Bass Werner
date: "Summer Full 2023"
output: html_document
---

## Database configuration and Set up

```{r connectToDatabase, eval = T, warning = F}
library (DBI)
library(RMySQL)

host <- "sql9.freemysqlhosting.net"
username <- "sql9638932"
password <- "r49v5qhMUF"
dbname <- "sql9638932"

dbcon <- dbConnect(
  MySQL(),
  dbname = dbname,
  host = host,
  user = username,
  password = password,
)
```

## Drop Existing Tables

```{sql drop_strikes, connection = dbcon, eval = T, echo = F}
DROP TABLE IF EXISTS strikes
```

```{sql drop_flights, connection = dbcon, eval = T, echo = F}
DROP TABLE IF EXISTS flights
```

```{sql drop_airports, connection = dbcon, eval = T, echo = F}
DROP TABLE IF EXISTS airports
```

```{sql drop_conditions, connection = dbcon, eval = T, echo = F}
DROP TABLE IF EXISTS conditions
```

### Create Airports Table

```{sql createAirportsTable, eval = T, connection=dbcon}
CREATE TABLE IF NOT EXISTS airports (
  aid INT PRIMARY KEY,
  airportName TEXT,
  airportState TEXT,
  airportCode TEXT
);
```

### Create Flights Table
  
```{sql createFlightsTable, eval = T, connection=dbcon}
CREATE TABLE IF NOT EXISTS flights (
  fid INT PRIMARY KEY,
  date DATE,
  origin INT,
  airline TEXT,
  aircraft TEXT,
  altitude INT,
  heavy BOOLEAN,
  FOREIGN KEY (origin) REFERENCES airports(aid)
);
```

### Create Conditions Table

```{sql CreateConditionsTable, eval = T, connection=dbcon}
CREATE TABLE IF NOT EXISTS conditions (
  cid INT PRIMARY KEY,
  sky_condition TEXT,
  explanation TEXT
);
```

### Create Strikes Table

```{sql CreateStrikesTable, eval = T, connection=dbcon}
CREATE TABLE IF NOT EXISTS strikes (
  sid INT PRIMARY KEY,
  fid INT,
  numbirds INT,
  impact TEXT,
  damage BOOLEAN,
  altitude INT CHECK (altitude >= 0),
  conditions INT,
  FOREIGN KEY (fid) REFERENCES flights(fid),
  FOREIGN KEY (conditions) REFERENCES conditions(cid)
);
```

## Manually Test Inserting Into Tables using sql code chunks

```{sql airportsTableInsertTest, connection=dbcon, eval=T}
INSERT INTO airports (aid, airportState, airportCode)
VALUES (1, 'California', 'LAX');
```

```{sql airportsTableOutputTest, connection=dbcon, eval=T}
SELECT * FROM airports;
```

```{sql flightsTableInsertTest, connection=dbcon, eval=T}
INSERT INTO flights (fid, date, origin, airline, aircraft, altitude, heavy)
VALUES (1, '2023-06-21', 1, 'Delta', 'Boeing 737', 35000, 1);
```

```{sql flightsTableOutputTest, connection=dbcon, eval=T}
SELECT * FROM flights;
```

```{sql conditionsTableInsertTest, connection=dbcon, eval=T}
INSERT INTO conditions (cid, sky_condition, explanation)
VALUES (1, 'Clear', 'No clouds in the sky.');
```

```{sql conditionsTableOutputTest, connection=dbcon, eval=T}
SELECT * FROM conditions;
```

```{sql strikesTableInsertTest, connection=dbcon, eval=T}
INSERT INTO strikes (sid, fid, numbirds, impact, damage, altitude, conditions)
VALUES (1, 1, 2, 'Engine failure', true, 30000, 1);
```

```{sql strikesTableOutputTest, connection=dbcon, eval=T}
SELECT * FROM strikes;
```

### Queries for deleting data from tables

```{sql strikesTableReset, connection=dbcon, eval=T}
DELETE FROM strikes;
```

```{sql flightsTableReset, connection=dbcon, eval=T}
DELETE FROM flights;
```

```{sql airportsTableReset, connection=dbcon, eval=T}
DELETE FROM airports;
```

```{sql conditionsTableReset, connection=dbcon, eval=T}
DELETE FROM conditions;
```

## Methods for sanitizing fields
```{r sanitizingMethods, eval=T, warning=FALSE}
# Parse date strings 
parseDateString <- function(date_strings) {
  parsed_dates <- sapply(date_strings, function(date_string) {
    if (date_string == "") {
      date <- "0000-00-00"
    } else {
      date_object <- strptime(date_string, format = "%m/%d/%Y %H:%M")
      date <- format(date_object, format = "%Y-%m-%d")
    }
    return(date)
  })
  
  return(parsed_dates)
}

# Method to sanitize empty fields or special characters
sanatizeStringForNullOrEmpty <- function(str) {
  #str <- str_escape(str)
  if (is.null(str) || str == "" || str == " ") {
    return("N/A")
  } else {
    return(str)
  }
}

# Method to parse Altitude 
parseAltitude <- function(altitude) {
  if (altitude == "") {
    altitude <- 0
  } else {
    altitude <- as.integer(gsub(",", "", altitude))
  }
  return(altitude)
}

# Method for parsing heavy flag
parseHeavy <- function(flag) {
  if (tolower(flag) == "yes") {
    return(1)
  } else {
    return(0)
  }
}

# Method for parsing damage
damageParsing <- function(damageString) {
  if (damageString == "Caused damage") {
    return(1)
  } else if (damageString == "No damage") {
    return(0)
  } else {
    return(-1)
  }
}

```

## Load CSV File into a data frame

```{r createDataframe, eval = T, warning = F}
bds.raw <- read.csv('BirdStrikesData.csv', 
                    header = T,
                   stringsAsFactors = F)
```

## Sanitize bds.raw
```{r}
# Sanitize empty or null strings in the airport column and replace them with "N/A"
bds.raw$airport <- sapply(bds.raw$airport, sanatizeStringForNullOrEmpty)

# Reformat and sanitize date string
bds.raw$flight_date <- sapply(bds.raw$flight_date, parseDateString)

# Parse all for empty or special characters 
bds.raw$origin <- sapply(bds.raw$origin, sanatizeStringForNullOrEmpty)
bds.raw$airline <- sapply(bds.raw$airline, sanatizeStringForNullOrEmpty)
bds.raw$aircraft <- sapply(bds.raw$aircraft, sanatizeStringForNullOrEmpty)

#Parse altitude
bds.raw$altitude_ft <- sapply(bds.raw$altitude_ft, parseAltitude)

#Parse heavy flag
bds.raw$heavy_flag <- sapply(bds.raw$heavy_flag, parseHeavy)

#Parse damage flag
bds.raw$damage <- sapply(bds.raw$damage, damageParsing)
```

## Populate the data frames

``` {r TESTpopulateDataFrames, eval = F, warning = F}
# Create data frames
airports.df <- data.frame(aid = integer(),
                          airportName = character(),
                          airportState = character(),
                          airportCode = character(),
                          stringsAsFactors = F)


# Helper function to check strings
get.airport.key <- function(airportName, airportState, df) {
  return(df$aid[which(df$airportName == airportName & df$airportState == airportState)])
}

# Initialize synthetic ID values
a.row = 1
f.row = 1
c.row = 1


  
    airports.df[a.row, "aid"] <- as.integer(9999999)
    airports.df[a.row, "airportName"] <- "Logan"
    airports.df[a.row, "airportState"] <- "MA"
    airports.df[a.row, "airportCode"] <- "N/A"
    a.row <- a.row + 1 # Increment p.row when a new genre is added

dbWriteTable(dbcon, "airports", airports.df, overwrite = F, append = T, row.names = FALSE)
# dbWriteTable(dbcon, "flights", flights.df, overwrite = T)
# dbWriteTable(dbcon, "conditions", conditions.df, overwrite = T)
# dbWriteTable(dbcon, "strikes", strikes.df, overwrite = T)
```

``` {r populateDataFrames, eval = T, warning = F}
# Create data frames
airports.df <- data.frame(aid = integer(),
                          airportName = character(),
                          airportState = character(),
                          airportCode = character(),
                          stringsAsFactors = F)

flights.df <- data.frame(fid = integer(),
                         date = character(),
                         origin = integer(),
                         airline = character(),
                         aircraft = character(),
                         altitude = integer(),
                         heavy = integer(),
                         stringsAsFactors = F)

conditions.df <- data.frame(cid = integer(),
                            sky_condition = character(),
                            explanation = character(),
                            stringsAsFactors = F)

strikes.df <- data.frame(sid = integer(),
                         fid = integer(),
                         numbirds = integer(),
                         impact = character(),
                         damage = integer(),
                         altitude = integer(),
                         conditions = character(),
                         stringsAsFactors = F)

# Helper function to check strings
get.airport.key <- function(airportName, airportState, df) {
  return(df$aid[which(df$airportName == airportName & df$airportState == airportState)])
}

# Helper function to check strings
get.condition.key <- function(sky_condition, df) {
  return(df$cid[which(df$sky_condition == sky_condition)])
}

# Helper function to check strings
get.flight.key <- function(date, aircraft, airline, origin, df) {
  return(df$fid[which(df$date == date & df$aircraft == aircraft 
                      & df$airline == airline & df$origin == origin)])
}

# Initialize synthetic ID values
a.row = 1
f.row = 1
c.row = 1

# Helper function to check if a value exists

for (row in 1:nrow(bds.raw)) {
  
  rid <- bds.raw[row, "rid"]
  airportName <- bds.raw[row, "airport"]
  aircraft <- bds.raw[row, "model"]
  numbirds <- bds.raw[row, "wildlife_struck"]
  impact <- bds.raw[row, "impact"]
  date <- bds.raw[row, "flight_date"]
  damage <- bds.raw[row, "flight_date"]
  airline <- bds.raw[row, "airline"]
  airportState <- bds.raw[row, "origin"]
  sky_condition <- bds.raw[row, "sky_conditions"]
  altitude <- bds.raw[row, "altitude_ft"]
  heavy <- bds.raw[row, "heavy_flag"]
  
  if (length(get.airport.key(airportName, airportState, airports.df)) == 0) {
    airports.df[a.row, "aid"] <- as.integer(a.row)
    airports.df[a.row, "airportName"] <- airportName
    airports.df[a.row, "airportState"] <- airportState
    airports.df[a.row, "airportCode"] <- "N/A"
    a.row <- a.row + 1 # Increment p.row when a new genre is added
  }
  
  if (length(get.flight.key(date, aircraft, airline, get.airport.key(airportName, airportState, airports.df), flights.df)) == 0) {
    flights.df[f.row, "fid"] <- as.integer(f.row)
    flights.df[f.row, "date"] <- date
    flights.df[f.row, "origin"] <- as.integer(get.airport.key(airportName, airportState, airports.df))
    flights.df[f.row, "airline"] <- airline
    flights.df[f.row, "aircraft"] <- aircraft
    flights.df[f.row, "heavy"] <- as.integer(heavy)
    flights.df[f.row, "altitude"] <- as.integer(altitude)
    f.row <- f.row + 1 # Increment p.row when a new genre is added
  }
  
  if (length(get.condition.key(sky_condition, conditions.df)) == 0) {
    conditions.df[c.row, "cid"] <- as.integer(c.row)
    conditions.df[c.row, "sky_condition"] <- sky_condition
    conditions.df[c.row, "explanation"] <- ""
    c.row <- c.row + 1 # Increment p.row when a new genre is added
  }
  
  strikes.df[row, "sid"] <- rid
  strikes.df[row, "fid"] <- as.integer(get.flight.key(date, aircraft, airline, get.airport.key(airportName, airportState, airports.df), flights.df))
  strikes.df[row, "numbirds"] <- as.integer(numbirds)
  strikes.df[row, "damage"] <- as.integer(damage)
  strikes.df[row, "altitude"] <- as.integer(altitude)
  strikes.df[row, "conditions"] <- as.integer(get.condition.key(sky_condition, conditions.df))
}

dbWriteTable(dbcon, "airports", airports.df, overwrite = F, append = T, row.names = FALSE)
dbWriteTable(dbcon, "flights", flights.df, overwrite = F, append = T, row.names = FALSE)
dbWriteTable(dbcon, "conditions", conditions.df, overwrite = F, append = T, row.names = FALSE)
dbWriteTable(dbcon, "strikes", strikes.df, overwrite = F, append = T, row.names = FALSE)
```

## Populate the flights data frame

```{sql airportsTableOutputTest, connection=dbcon, eval=T}
SELECT * FROM airports;
```

```{sql flightsTableOutputTest, connection=dbcon, eval=T}
SELECT * FROM flights;
```

```{sql conditionsTableOutputTest, connection=dbcon, eval=T}
SELECT * FROM conditions;
```

```{sql strikesTableOutputTest, connection=dbcon, eval=T}
SELECT * FROM strikes;
```

```{sql strikesTableOutputTest, connection=dbcon, eval=T}
SELECT COUNT(*) FROM strikes AS total_strikes;
```

## Find the top 10 states with the greatest number of bird strike incidents

```{sql question8, connection=dbcon, eval=T}
SELECT airportState AS state, COUNT(*) AS incidents
FROM strikes
JOIN flights ON strikes.fid = flights.fid
JOIN airports ON flights.origin = airports.aid
GROUP BY airportState
ORDER BY incidents DESC
LIMIT 10;
```

## Airlines with an above average number of bird strike incidents

```{sql question9, connection=dbcon, eval=T}
SELECT airline, COUNT(*) AS incidents
FROM strikes
JOIN flights ON strikes.fid = flights.fid
JOIN airports ON flights.origin = airports.aid
GROUP BY airline
HAVING incidents > (SELECT AVG(incident_count) FROM (SELECT COUNT(*) AS incident_count FROM strikes JOIN flights ON strikes.fid = flights.fid GROUP BY airline) AS avg_counts)
ORDER BY incidents DESC;
```

```{sql connection=dbcon}
SELECT AVG(incident_count) FROM (SELECT COUNT(*) AS incident_count FROM strikes JOIN flights ON strikes.fid = flights.fid GROUP BY airline) AS avg_counts
```

## Total number of birds struck by aircraft by month 

```{r question10, eval = T, warning = F}

query <- "SELECT DATE_FORMAT(date, '%m') AS month, SUM(numbirds) AS total_birds_struck_aircraft
  FROM strikes
  JOIN flights ON strikes.fid = flights.fid
  WHERE DATE_FORMAT(date, '%m') <> '00'  -- Exclude rows with month '00'
  GROUP BY month
  ORDER BY month;"

result <- dbGetQuery(dbcon, query)

df <- as.data.frame(result)

print(df)
```

## Data Graph For Total Number of Birds Struck by Aircraft by Month 

```{r question11, eval = T, warning = F}
# Create a column chart
barplot(df$total_birds, names.arg = df$month, xlab = "Month", ylab = "Number of Birds",
        main = "Bird Strikes per Month", col = "blue", ylim = c(0, max(df$total_birds)))

# Add data labels
text(x = 1:length(df$total_birds), y = df$total_birds, labels = df$total_birds, pos = 3, cex = 0.8)

# Add a legend
legend("topright", legend = c("Number of Birds"), fill = "blue")

# Adjust the plot margins
par(mar = c(5, 4, 4, 2) + 0.1)
```

## Stored Procedure for Question 12

```{sql question12reset, connection=dbcon, eval=T}
DROP PROCEDURE IF EXISTS AddNewStrike;
```

```{sql question12, connection=dbcon, eval=T}
CREATE PROCEDURE AddNewStrike(
  IN fid INT,
  IN numbirds INT,
  IN impact TEXT,
  IN damage BOOLEAN,
  IN altitude INT,
  IN cid INT
)
BEGIN

  INSERT INTO strikes (fid, numbirds, impact, damage, altitude, conditions)
  VALUES (fid, numbirds, impact, damage, altitude, cid);
END;
```

```{r question12results, eval = T, warning = F}
# Call stored procedure
query <- "CALL AddNewStrike(
  1,
  10,
  'wing',
  0,
  5000,
  1
)"

# Envoke stored procedure
dbGetQuery(dbcon, query)

check <- "SELECT *
FROM strikes
ORDER BY sid DESC
LIMIT 1;"

result <- dbGetQuery(dbcon, check)

# Print the row that would have been recently added
print(result)
```

## Disconnect From Database

```{r}
dbDisconnect(dbcon)
```